Retrofitting Effect Handlers onto OCaml

### 2020-11-11 retro-concurrency.tex
Minor rewords and tweaks: sections 1-3

\begin{abstract}
  Effect handlers have gained immense interest in the recent years as a means
  for modular programming with user-defined effects. The runtime semantics of
  effect handlers is particularly attractive as they allow advanced control
  flow mechanisms such as first-class continuations, generators, async/await,
  coroutines to be composably expressed. In this paper, we present the design,
  a full-fledged implementation and evaluation of effect handlers for OCaml, an
  industrial-strength, multi-paradigm programming language. OCaml tends to be
  widely used for systems programming, and is compatible with program analysis
  tools that inspect the stack such as debuggers and profilers. Retrofitting
  effect handlers onto OCaml is challenging since OCaml does not have any
  non-local control flow mechanisms other than exceptions. Our implementation
  of effect handlers for OCaml imposes negligible overhead on code that does
  not use effect handlers, and remains compatible with program analysis tools
  that inspect the stack.
\end{abstract}

@@

On line 4, we declare an effect |DivBy0| which is parameterised by an integer,
which when performed returns result of division. If the denominator is 0, then
we perform |DivBy0| effect with the numerator value (line 12) so that the
context may decide the return value. We can get the |iOS| semantics by wrapping
the evaluator computation with the following handler:
\begin{lstlisting}
let ios e = match eval e with
| v -> v
| effect (DivBy0 _) k -> continue k Error
\end{lstlisting}
If the evaluator returns a value, we simply return that. In case the evaluator
performs the |DivBy0| effect, we catch the effect in the effect case along with
the continuation |k| of the corresponding |perform| \emph{delimited} by this
handler. In the case of |iOS|, we |continue| the suspended computation with
|Error| value. Hence, the computation |ios(1/0)| returns |Error|. On the
other hand, we can get |gsearch| semantics by enclosing the evaluator in the
following handler:
\begin{lstlisting}
let gsearch e = match eval e with
| v -> v
| effect (DivBy0 v) k ->
  if v = 0 then continue k Error
  else if v > 0 then continue k (Inf P)
  else continue k (Inf N)
\end{lstlisting}
The computation |gsearch(1/0)| returns positive infinity (|Inf P|).

@@

We can extend the calculator modularly with the ability to read user input
using the |Read| effect defined earlier as follows:
\begin{lstlisting}
type exp = ... | Inp
effect Read : in_channel -> string
let rec eval e = match e with
| ...
| Inp -> int_of_string(perform(Read stdin))
\end{lstlisting}
The |Read| effect can then be handled by another handler enclosing the
division-by-zero handler:
\begin{lstlisting}
let run e = match gsearch e with
| v -> v
| effect (Read chan) k ->
  register_async_input_line (chan, k);
  continue (get_ui_k ()) ()
\end{lstlisting}
\noindent which registers the read operation to be performed asynchronously
with the continuation as the callback. Subsequently, it resumes the
user-interface continuation so that the application remains responsive. Observe
that neither of the handlers needs to be aware of the other.

@@

There are many research languages and libraries built around effect
handlers~\cite{Leijen14,Hillerstrom20,Pyro,Frank,Eff}. Unlike these efforts, our goal is to
retrofit effect handlers onto the OCaml programming language, which has been in
continuous use for the past 25 years in large codebases including verification
tools~\cite{everest,Coq}, mission critical software systems~\cite{astree} and
latency sensitive networked applications~\cite{Madhavapeddy13}. OCaml is
particularly favoured for its competitive yet predictable performance, with
fast foreign-function interface (FFI) and excellent compatibility with program
analysis tools such as debuggers and profilers that utilise DWARF stack unwind
tables~\cite{DWARF} to obtain a backtrace. Our primary motivation for adding
effect handlers to OCaml is to support (1) highly concurrent applications such
as web servers to be expressed naturally in \emph{direct-style} as opposed to
using callbacks and (2) efficient nested parallel programs in the vein of
Cilk~\cite{Cilk}, OpenMP~\cite{OpenMP}, X10~\cite{Charles05}, Intel
TBB~\cite{IntelTBB}, and variants of Parallel ML~\cite{MaPLe, Fluet10,
Sivaramakrishnan14}.

OCaml, however, does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. One standard way of implementing continuations
is to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Leijen14}. If the compiler were to use CPS, then reifying
continuations in the source language is relatively straight-forward. But OCaml
does not use CPS IR, and changing the OCaml compiler to utilise CPS IR would be
an enormous undertaking. CPS IR would also affect the performance profile of
OCaml applications due to increased memory allocations as continuation closures
get allocated on the heap~\cite{Farvardin20}. Moreover, with CPS, stack
unwinding is meaningless and hence, we would lose compatibility with tools that
inspect the program stack. Hence, we choose not to use CPS translation and
represent continuations as call stacks.

@@

The need to host millions of continuations at the same time rules out the use
of large contiguous stack space as in C for continuations. Instead, we resort
to using small initial stacks and growing the stacks on demand. As a result,
OCaml functions, irrespective of whether they use effect handlers, need to
perform stack overflow checks and external C functions, which do not have stack
overflow checks, must be performed on a separate system stack. Additionally, we
must generate DWARF stack unwind tables for stacks that may be non-contiguous.
In this work, we develop the compiler and runtime support for implementing
efficient effect handlers for OCaml that satisfy these requirements.

Our work is also timely. WebAssembly~\cite{Hass17} community group is actively
considering effect handlers as one of the mechanism for supporting concurrency,
asynchronous I/O and generators~\cite{WasmProposal}. Project Loom~\cite{loom} is an
OpenJDK project which adds virtual threads and delimited continuations to Java.
Swift concurrency roadmap~\cite{swift} includes direct-style asynchronous
programming and structured concurrency as milestones. We believe that our
design choices will inform similar choices to be made in the other
industrial-strength languages.

@@

\begin{itemize}
  \item the design and implementation of one-shot effect handlers for OCaml.
    Our design retains OCaml's compatibility with program analysis tools that
    inspect the stack using DWARF unwind tables. We have validated our DWARF
    unwind tables using an automated validator tool~\cite{Bastian19}.
  \item a formal operational semantics for the effect handler implementation in
    OCaml.
	\item extensive evaluation which shows that our implementation has negligible
		impact on code that does not use effect handlers, and serves as an
		efficient foundation for scalable concurrent programming.
\end{itemize}

@@

In order to illustrate the layout of the stock OCaml stack, consider the
program in Figures~\ref{code:meander_ml} and~\ref{code:meander_c}. The OCaml
main function |omain| installs two exception handlers |h1| and |h2| to handle
exceptions |E1| and |E2|. |omain| calls the external C function |ocaml_to_c|,
which in turn calls back into OCaml function |c_to_ocaml|, which raises the
exception |E1|. OCaml supports raising exceptions in C as well as throwing
exceptions across external calls. Hence, the exception |E1| gets caught in the
handler |h1|, and |omain| returns |42|. The layout of the stack in native code
backend just before raising the exception in |c_to_ocaml| is illustrated in
Figure~\ref{fig:stock_stack}. Note that the stack grows downwards.

OCaml uses the same hardware stack as C, and hence the program stack has
alternating sequences of C and OCaml frames. However, unlike C, OCaml does not
create pointers into OCaml frames. OCaml uses the hardware support for |call|
and |return| instructions for function calls and returns. OCaml does not
perform explicit stack overflow checks in code, and, just like C, relies on the
guard page at the end of the stack region to detect stack overflow. Stack
overflow is detected by a memory fault, and |Stack_overflow| exception is
raised to unwind the stack.

@@

OCaml does not use the C calling convention. In particular, there are no
callee-saved registers in OCaml. On the |x86-64| backend, the C callee-saved
registers  |r15| holds the \emph{allocation pointer} into the minor heap used
for bump pointer allocation, and |r14| holds a reference to the |Caml_state|
table of global variables used by the runtime. This makes external calls
extremely fast in OCaml. If the external function does not allocate in the
OCaml heap, then the external functions can be called directly, and no
bookkeeping is necessary. For external functions which allocate in the OCaml
heap, the cached allocation pointer is saved to a field in the |Caml_state|
before the external call, and loaded back on return. Similarly, callbacks into
OCaml are also cheap. Callbacks into OCaml involves loading the arguments in
the right registers and calling the OCaml function. OCaml callbacks are
relatively common as the garbage collector (GC) finalisation functions are
executed as OCaml callbacks.

@@

The lack of callee-saved registers also makes exception handling fast as there
are no callee-saved registers to be restored when the exception handler runs.
Hence, installing an exception handler simply pushes the program counter (|pc|)
of the handler and the current exception pointer (|exn_ptr| -- a field in
|Caml_state|). At this point, the exception pointer is updated to be the
current stack pointer (|rsp|). This creates a linked-list of exception handler
frames on the stack as shown in Figure~\ref{fig:stock_stack}. Raising an
exception simply sets |rsp| to |exn_ptr|, loads the saved |exn_ptr|, and jumps
to the |pc| of the handler.

@@

In order to forward exceptions across C frames, the C stub function
|caml_call_ocaml|, which is the entry point for both the main OCaml code and
callbacks into OCaml, pushes an exception handler frame that forwards the
exception to the innermost OCaml exception handler (|raise_exn_c| in
Figure~\ref{fig:stock_stack}), or prints a fatal error (|fatal_uncaught|) if
there are no enclosing handlers. Exceptions are so cheap in OCaml that it is
very common to use them for \emph{local} control flow.

@@

OCaml generates \emph{stack maps} in order to accurately identify roots on the
stack for assisting GC. For every call point in the program, OCaml compiler
emits the size of the frame and the set of all live registers in the frame
containing pointers into the heap. During a GC, the OCaml stack is walked and
the roots are marked skipping over the C frames.

OCaml also generates precise DWARF unwind information for OCaml, thanks to
which debuggers such as |gdb| and |lldb|, and profilers such as |perf| work
out-of-the-box. For example, for the program in Figures~\ref{code:meander_ml}
and~\ref{code:meander_c}, one could set a break point in |gdb| at
|caml_raise_exn| to get the backtrace in Figuere~\ref{code:gdb_backtrace} which
corresponds to the stack layout in Figure~\ref{fig:stock_stack}.

@@

In this section, we refine the effect handler design for the OCaml use case.
The primary motivation for adding effect handlers to OCaml is to support
non-local control flow mechanism to implement asynchronous I/O in direct style,
generators and enable task parallelism. These use cases do not resume the
continuations more than once. Hence, our continuations are one-shot, and
resuming the continuation more than once raises |Invalid_argument| exception.
It is well-known that one-shot continuations can be implemented
efficiently~\cite{Bruggeman96}.

@@

The interaction of non-local control flow with systems programming is quite
subtle~\cite{TFP17}. Consider the following function that uses blocking I/O
functions from the OCaml standard library to copy data from the input channel
|ic| to the output channel |oc|:
\begin{lstlisting}
let rec copy ic oc =
  try output_string oc ((input_line ic) ^ "\n");
      copy ic oc
  with End_of_file -> close_in ic; close_out oc
	   | e -> close_in ic; close_out oc; raise e
\end{lstlisting}
The function |input_line| raises |End_of_file| exception on reaching the end of
input, which is handled by the exception handler which closes the channels. The
|close_*| functions do nothing if the channel is already closed. The code is
written in a defensive style to handle other exceptional cases such as the
channels being closed externally. Both |input_line| and |output_string| raise
|Sys_error| exception if the channel is closed. In this case, the channels are
closed and the exception is reraised to communicate the exceptional behaviour
to the caller.

One of our goals(\S\ref{sec:req}) is to make this code transparently
asynchronous. We can define effects for performing the I/O operations  and wrap
them up in functions with the same signature as the one from the standard
library:
\begin{lstlisting}
effect In_line : in_channel -> string
effect Out_str : out_channel -> string -> unit
let input_line ic = perform (In_line ic)
let output_string oc = perform (Out_str oc)
\end{lstlisting}

@@

Program that use resources such as channels are usually written defensively
with the assumption that calling a function will return \emph{exactly once},
either normally or exceptionally. Since effect handlers in Multicore OCaml do
not ensure that all the effects are handled, if the function performs an effect
with no matching handler, then the function \emph{will not return at all}. To
remedy this, when such an effect bubbles up to the top-level, we discontinue
the continuation with |Unhandled| exception so that the exception handlers may
run and clean up the resources.

>>>>

\begin{abstract}
  Effect handlers have gained immense interest in the recent years as a means
for modular programming with user-defined effects. The runtime semantics of
effect handlers is particularly attractive as they allow advanced control flow
mechanisms such as first-class continuations, generators, async/await,
coroutines to be composably expressed. This paper presents a design, a
full-fledged implementation and an evaluation of effect handlers for OCaml.
OCaml is industrial-strength multi-paradigm programming language that tends to
be widely used for systems programming, our implementation strives to maintain
the existing performance profile and toolchain compatibility. Retrofitting
effect handlers onto OCaml is challenging since OCaml does not have any
non-local control flow mechanisms other than exceptions. Our implementation of
effect handlers for OCaml imposes negligible overhead on code that does not use
effect handlers, remains compatible with program analysis tools that inspect the
stack and is efficient for code that makes use of effect handlers.
\end{abstract}

@@

On line 4, we declare an effect |DivBy0| that is parameterised by an integer,
which when performed returns the result of division-by-zero. If the denominator
is 0, then we perform |DivBy0| effect with the numerator value (line 12) so that
the context may decide the return value. We can get the |iOS| semantics by
wrapping the evaluator with the following handler:
\begin{lstlisting}
let ios e = match eval e with
| v -> v
| effect (DivBy0 _) k -> continue k Error
\end{lstlisting}
If the evaluator returns a value, we simply return that. If the evaluator
performs the |DivBy0| effect, we catch the effect in the effect case along with
the continuation |k| of the corresponding |perform| \emph{delimited} by this
handler. In the case of |iOS|, we |continue| the suspended computation with
|Error| value. Hence, the computation |ios(1/0)| returns |Error|. On the
other hand, we can get the |gsearch| semantics by enclosing the evaluator in the
following handler:
\begin{lstlisting}
let gsearch e = match eval e with
| v -> v
| effect (DivBy0 v) k ->
  if v = 0 then continue k Error
  else if v > 0 then continue k (Inf P)
  else continue k (Inf N)
\end{lstlisting}
The computation |gsearch(1/0)| returns positive infinity (|Inf P|).

@@

We can modularly extend the calculator to read user input using the |Read|
effect defined earlier as follows:
\begin{lstlisting}
type exp = ... | Inp
effect Read : in_channel -> string
let rec eval e = match e with
| ...
| Inp -> int_of_string(perform(Read stdin))
\end{lstlisting}
The |Read| effect can then be handled by another handler enclosing the
division-by-zero handler:
\begin{lstlisting}
let run e = match gsearch e with
| v -> v
| effect (Read chan) k ->
  register_async_input_line (chan, k);
  continue (get_ui_k ()) ()
\end{lstlisting}
\noindent which registers the read operation to be performed asynchronously
with the continuation as the callback. Subsequently, it resumes the
user-interface continuation so that the application remains responsive. Observe
that neither of the handlers needs to be aware of the other.

@@

There are many research languages and libraries built around effect
handlers~\cite{Leijen14,Hillerstrom20,Pyro,Frank,Eff}. Unlike these efforts, our
goal is to retrofit effect handlers onto the OCaml programming language, which
has been in continuous use for the past 25 years in large codebases including
verification tools~\cite{everest,Coq}, mission critical software
systems~\cite{astree} and latency sensitive networked
applications~\cite{Madhavapeddy13}. OCaml is particularly favoured for its
competitive yet predictable performance, with fast foreign-function interface
(FFI) and excellent compatibility with program analysis tools such as debuggers
and profilers that utilise DWARF stack unwind tables~\cite{DWARF} to obtain a
backtrace. Our primary motivation for adding effect handlers to OCaml is to
support (1) highly concurrent applications such as web servers using a
natural \emph{direct-style} as opposed to using callbacks and (2) efficient
nested parallel programs in the vein of Cilk~\cite{Cilk}, OpenMP~\cite{OpenMP},
X10~\cite{Charles05}, Intel TBB~\cite{IntelTBB}, and variants of Parallel
ML~\cite{MaPLe, Fluet10, Sivaramakrishnan14}.

OCaml, however, does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. One standard way of implementing continuations is
to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Leijen14}. If the compiler were to use CPS, then
reifying continuations in the source language is relatively straight-forward.
But OCaml does not use CPS IR, and changing the OCaml compiler to utilise CPS IR
would be an enormous undertaking. Utilizing a CPS IR would also affect the
performance profile of OCaml applications due to increased memory allocations as
continuation closures get allocated on the heap~\cite{Farvardin20}. Moreover,
with CPS, stack unwinding is meaningless and hence, we would lose compatibility
with tools that inspect the program stack. Hence, we choose not to use CPS
translation and represent continuations as call stacks.

@@

The need to host millions of continuations at the same time rules out the use
of a large contiguous stack space as in C for continuations. Instead, we resort
to using small initial stacks and growing the stacks on demand. As a result,
OCaml functions, irrespective of whether they use effect handlers, need to
perform stack overflow checks and external C functions, which do not have stack
overflow checks, must be performed on a separate system stack. Additionally, we
must generate DWARF stack unwind tables for stacks that may be non-contiguous.
In this work, we develop the compiler and runtime support for implementing
efficient effect handlers for OCaml that satisfy these requirements.

Our work is also timely. The WebAssembly~\cite{Hass17} community group is
actively considering effect handlers as one of the mechanism for supporting
concurrency, asynchronous I/O and generators~\cite{WasmProposal}. Project
Loom~\cite{loom} is an OpenJDK project which adds virtual threads and delimited
continuations to Java. Swift concurrency roadmap~\cite{swift} includes
direct-style asynchronous programming and structured concurrency as milestones.
We believe that our design choices will inform similar choices to be made in the
other industrial-strength languages.

@@

\begin{itemize}
  \item the design and implementation of one-shot effect handlers for OCaml.
    Our design retains OCaml's compatibility with program analysis tools that
    inspect the stack using DWARF unwind tables. We have validated our DWARF
    unwind tables with the assistance of an automated validator
    tool~\cite{Bastian19}.
  \item a formal operational semantics for the effect handler implementation in
    OCaml.
	\item extensive evaluation which shows that our implementation has negligible
		impact on code that does not use effect handlers, and serves as an
		efficient foundation for scalable concurrent programming.
\end{itemize}

@@

Consider the layout of the stock OCaml stack for the
program in Figures~\ref{code:meander_ml} and~\ref{code:meander_c}. The OCaml
main function |omain| installs two exception handlers |h1| and |h2| to handle
the exceptions |E1| and |E2|. |omain| calls the external C function |ocaml_to_c|,
which in turn calls back into the OCaml function |c_to_ocaml|, which raises the
exception |E1|. OCaml supports raising exceptions in C as well as throwing
exceptions across external calls. Hence, the exception |E1| gets caught in the
handler |h1|, and |omain| returns |42|. The layout of the stack in the native code
backend just before raising the exception in |c_to_ocaml| is illustrated in
Figure~\ref{fig:stock_stack}. Note that the stack grows downwards.

OCaml uses the same program stack as C, and hence the stack has
alternating sequences of C and OCaml frames. However, unlike C, OCaml does not
create pointers into OCaml frames. OCaml uses the hardware support for |call|
and |return| instructions for function calls and returns. OCaml does not
perform explicit stack overflow checks in code, and, just like C, relies on the
guard page at the end of the stack region to detect stack overflow. Stack
overflow is detected by a memory fault and a |Stack_overflow| exception is
raised to unwind the stack.

@@

OCaml does not use the C calling convention. In particular, there are no
callee-saved registers in OCaml. On the |x86-64| backend, the OCaml runtime
makes use of C callee-saved registers: |r15| holds the \emph{allocation pointer}
into the minor heap used for bump pointer allocation, and |r14| holds a
reference to the |Caml_state| table of global variables used by the runtime.
This makes external calls extremely fast in OCaml. If the external function does
not allocate in the OCaml heap, then it can be called directly and no
bookkeeping is necessary. For external functions which allocate in the OCaml
heap, the cached allocation pointer is saved to |Caml_state| before the external
call and it is restored on return. Similarly, callbacks into OCaml are also
cheap: these involve loading the arguments in the right registers and calling
the OCaml function. OCaml callbacks are relatively common as the garbage
collector (GC) finalisation functions are executed as OCaml callbacks.

@@

The lack of callee-saved registers also makes exception handling fast as there
are no callee-saved registers to be restored when the exception handler runs.
Installing an exception handler simply pushes the program counter (|pc|) of the
handler and the current exception pointer (|exn_ptr| -- a field in |Caml_state|)
to the stack. At this point, the current exception pointer is updated to be the
current stack pointer (|rsp|). This creates a linked-list of exception handler
frames on the stack as shown in Figure~\ref{fig:stock_stack}. Raising an
exception simply sets |rsp| to |exn_ptr|, loads the saved |exn_ptr|, and jumps
to the |pc| of the handler.

@@

In order to forward exceptions across C frames, the C stub function
|caml_call_ocaml|, pushes an exception handler frame that forwards the
exception to the innermost OCaml exception handler (|raise_exn_c| in
Figure~\ref{fig:stock_stack}) or prints a fatal error (|fatal_uncaught|) if
there are no enclosing handlers. Exceptions are so cheap in OCaml that it is
common to use them for \emph{local} control flow.

@@

OCaml generates \emph{stack maps} in order to accurately identify roots on the
stack for assisting the GC. For every call point in the program, the OCaml compiler
emits the size of the frame and the set of all live registers in the frame
containing pointers into the heap. During a GC, the OCaml stack is walked and
the roots are marked skipping over the C frames.

OCaml also generates precise DWARF unwind information for OCaml, thanks to
which debuggers such as |gdb| and |lldb|, and profilers such as |perf| work
out-of-the-box. For example, for the program in Figures~\ref{code:meander_ml}
and~\ref{code:meander_c}, one could set a break point in |gdb| at
|caml_raise_exn| to get the backtrace in Figure~\ref{code:gdb_backtrace} which
corresponds to the stack in Figure~\ref{fig:stock_stack}.

@@

In this section, we refine the effect handler design for the OCaml use case. The
primary motivation for adding effect handlers to OCaml is to support a non-local
control flow mechanism to implement asynchronous I/O in direct style, generators
and enable task parallelism. These use cases do not resume the continuations
more than once. Hence, our continuations are one-shot, and resuming the
continuation more than once raises |Invalid_argument| exception. It is
well-known that one-shot continuations can be implemented
efficiently~\cite{Bruggeman96}.

@@

The interaction of non-local control flow with systems programming is quite
subtle~\cite{TFP17}. Consider the following function that uses blocking I/O
functions from the OCaml standard library to copy data from the input channel
|ic| to the output channel |oc|:
\begin{lstlisting}
let rec copy ic oc =
  try output_string oc ((input_line ic) ^ "\n");
      copy ic oc
  with End_of_file -> close_in ic; close_out oc
	   | e -> close_in ic; close_out oc; raise e
\end{lstlisting}
The function |input_line| raises an |End_of_file| exception on reaching the end of
input, which is handled by the exception handler which closes the channels. The
|close_*| functions do nothing if the channel is already closed. The code is
written in a defensive style to handle other exceptional cases such as the
channels being closed externally. Both |input_line| and |output_string| raise
|Sys_error| exception if the channel is closed. In this case, the channels are
closed and the exception is reraised to communicate the exceptional behaviour
to the caller.

One of our goals (\S\ref{sec:req}) is to make this code transparently
asynchronous. We can define effects for performing the I/O operations and wrap
them up in functions with the same signature as the one from the standard
library:
\begin{lstlisting}
effect In_line : in_channel -> string
effect Out_str : out_channel -> string -> unit
let input_line ic = perform (In_line ic)
let output_string oc = perform (Out_str oc)
\end{lstlisting}

@@

Programs that use resources such as channels are usually written defensively
with the assumption that calling a function will return \emph{exactly once},
either normally or exceptionally. Since effect handlers in Multicore OCaml do
not ensure that all the effects are handled, if the function performs an effect
with no matching handler, then the function \emph{will not return at all}. To
remedy this, when such an effect bubbles up to the top-level, we discontinue the
continuation with an |Unhandled| exception so that the exception handlers may
run and clean up the resources.
